(Deep) Induction Rules for GADTs

### 2021-09-09 induction-paper/acm-rev-ind-paper.tex
added cpp ind paper

\documentclass[9pt]{entcs}
\usepackage{entcsmacro}
\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}
\usepackage{amscd}
\usepackage{MnSymbol}
\usepackage{xcolor}
\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
\usepackage{stmaryrd}
\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usepackage{bussproofs}
\EnableBpAbbreviations

\sloppy
% The following is enclosed to allow easy detection of differences in
% ascii coding.
% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
% Digits        0 1 2 3 4 5 6 7 8 9
% Exclamation   !           Double quote "          Hash (number) #
% Dollar        $           Percent      %          Ampersand     &
% Acute accent  '           Left paren   (          Right paren   )
% Asterisk      *           Plus         +          Comma         ,
% Minus         -           Point        .          Solidus       /
% Colon         :           Semicolon    ;          Less than     <
% Equals        =3D           Greater than >          Question mark ?
% At            @           Left bracket [          Backslash     \
% Right bracket ]           Circumflex   ^          Underscore    _
% Grave accent  `           Left brace   {          Vertical bar  |
% Right brace   }           Tilde        ~

\bibliographystyle{entcs}

\def\lastname{Johann and Ghiorzi}

\input{macros}

\begin{document}
\begin{frontmatter}
  \vspace*{-0.1in}
  \title{(Deep) Induction Rules for GADTs\vspace*{-0.1in}}
  \author{Patricia Johann~~~~}%\thanksref{JohannEmail}}
  \author{Enrico Ghiorzi~~~~}%\thanksref{GhiorziEmail}}
%  \author{Daniel Jeffries}%\thanksref{JeffriesEmail}}
%  \thanks[JohannEmail]{Email: \href{mailto:johannp@appstate.edu} {\texttt{\normalshape johannp@appstate.edu}}}
%  \thanks[GhiorziEmail]{Email: \href{mailto:ghiorzie@appstate.edu} {\texttt{\normalshape ghiorzie@appstate.edu}}}
%  \thanks[JeffriesEmail]{Email: \href{mailto:jeffriesd@appstate.edu} {\texttt{\normalshape jeffriesd@appstate.edu}}}
  \address{$\mathtt{\{johannp,ghiorzie\}@appstate.edu}$\\
    Appalachian State University}

\begin{abstract} 
  Deep data types are those that are contructed from other data types,
  including, possibly, themselves.  In this case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction can be used to prove properties of
  nested types, including truly nested types, that cannot be proved
  via structural induction.  In this paper we show how to extend deep
  induction to GADTs that are not truly nested GADTs. We also show
  that the techniques developed in this paper do not suffice for
  extending deep induction to truly nested GADTs.\looseness=-1
\end{abstract}

\end{frontmatter}

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep
ADTs~\cite{jp20}.\footnote{An ADT/nested type/GADT is {\em deep} if it
  is (possibly mutually inductively) defined in terms of other
  ADTs/nested types/GADTs (including, possibly, itself).}  The
following data type of rose trees, here coded in Agda and defined in
terms of the standard type $\mathsf{List}$ of lists (see
Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:

\[\begin{array}{l}
\mathsf{data\, Rose\, : Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;empty\, :\,\forall \{A : Set\} \to Rose\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
\end{array}\]
\noindent
The induction rule Coq automatically generates for (the analogous Coq
definition of) rose trees is

\vspace*{-0.075in}

\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
  \forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains.  With data types becoming ever more advanced, and with
deeply structured such types becoming increasingly ubiquitous in
formalizations, such an enhanced notion of induction is essential if
proof assistants are to be
%it is critically important that proof assistants
%%{\color{red} (e.g., Coq, Agda, Lean)}
%are
able to automatically generate genuinely useful induction rules for
data types that go
%well
beyond traditional ADTs. These
%Such data types
include not just deep ADTs, but also (truly\footnote{A truly nested
  type is a nested type that is defined over itself. The data type
  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
  concrete example.}) nested types~\cite{bm98},
%, such as $\mathsf{Rose}$ above, as well as truly nested types
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{dyb94}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to deliver induction rules
appropriate to nested types, including ADTs.  For the (deep) ADT of
rose trees, for example, it gives the following genuinely useful
induction rule:

\vspace*{-0.075in}

\begin{equation}\label{eq:rose}
\begin{array}{l}
\mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
  \to P\,empty \to}\\ 
\mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
  List\,(Rose\,A))\to Q\,a \to List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
\mathsf{\hspace*{0.15in}\forall \,(x :
  Rose\,A) \to Rose^\land\,A\,Q\, x \to P\,x} 
\end{array}
\end{equation}

\noindent
%Here, $\mathsf{List^\land}$ lifts its predicate argument $\mathsf{P}$
%on data of type $\mathsf{Rose\,a}$ to a predicate on data of type
%$\mathsf{List\,(Rose\,a)}$ asserting that $\mathsf{P}$ holds for every
%element of its argument list, and similarly for the lifting
%$\mathsf{Rose^\land}$ and the predicate $\mathsf{Q}$.
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds} below. The predicate lifting for
  a container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$
  simply traverses containers of that type and applies its predicate
  argument pointwise to the constituent data of type $\mathsf{A}$.
  The ability to define predicate liftings for more general data types
  will be critical to deriving their deep induction rules in
  Section~\ref{sec:framework}.} Deep induction was also shown
in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and (truly) nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds} below.

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of data whose types are pair-structured, rather
than sequences of arbitrary type, as does $\mathsf{const}$.

If one or more of the constructors for a GADT $\mathsf{G}$ return
structured instances of $\mathsf{G}$, then the GADT will have two
distinct, but equally natural, semantics: a functorial semantics
interpreting it as a left Kan extension~\cite{mac71}, and a parametric
semantics interpreting it as the interpretation of its Church
encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
difference in the two semantics is that the former views GADTs as
their {\em functorial completions}~\cite{jp19}, and thus as containing
more data than just those expressible in syntax. By contrast, the
latter views them as what might be called {\em syntax-only}
GADTs. Happily, these two views of GADTs coincide for those GADTs that
are ADTs or (other, including truly) nested types.  However, both they
and their attendant properties differ greatly for proper GADTs. In
fact, the functorial and parametric semantics for proper GADTs are
sufficiently disparate that, by contrast with the semantics
customarily given for ADTs and nested
types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
define a functorial parametric semantics for GADTs~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so GADTs viewed as their functorial completions
do not obviously support induction rules.  This makes sense
intuitively: induction is a syntactic proof technique, so it may not
be possible to use it to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs. Indeed, this
paper gives the first-ever deep induction rules
%--- deep or otherwise ---
for proper GADTs.
%It also shows how the structural induction rule for
%a GADT can be derived as an instance of its deep induction rule.
But it actually delivers far more: it gives a {\em general framework}
for deriving deep induction rules for GADTs that can be instantiated
to particular such data type of interest. This framework can serve as
a basis for extending modern proof assistants' automatic generation of
structural induction rules for ADTs to automatic generation of deep
induction rules for GADTs. In addition, as for ADTs and nested types,
the structural induction rule for any GADT can be recovered from its
deep induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

@@

A structural induction rule for a data type allows us to prove that if
a predicate holds for every element inductively produced by the data
type's constructors then it holds for every element of the data type.
In this paper, we are interested in induction rules for proof-relevant
predicates.  A proof-relevant predicate on $\mathsf{A : Set}$ is a
function $\mathsf{P\,:\,A \to Set}$ mapping each $\mathsf{a : A}$ to
the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
structural induction rule for the list type

\vspace*{-0.2in}

\begin{equation*}\label{eq:list}
\begin{array}{l}
\mathsf{data\ List : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
\mathsf{\;\;\;\;\;\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
\end{array}
\end{equation*}

@@

\noindent
is

\vspace*{-0.3in}

\begin{equation*}
\mathsf{
\forall (A : Set) (P : List\,A \to Set)
\to P\,nil
\to \big( \forall (a : A) (as: List\,A)
\to P\,as
\to P\,(cons\,a\,as)\big)
\to \forall (as : List\,A)
\to P\, as
}
\end{equation*}

>>>>

\documentclass[sigplan,10pt,anonymous,review]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{graphicx}
\usepackage{adjustbox}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}
\usepackage{ccicons}
\usepackage{verbatim}
\usepackage{amscd}
%\usepackage{MnSymbol}
\usepackage{xcolor}
\usepackage{bbold}
\usepackage{url}
\usepackage{upgreek}
\usepackage{stmaryrd}
\usepackage{lipsum}
\usepackage{tikz-cd}
\usetikzlibrary{cd}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usepackage{bussproofs}
\EnableBpAbbreviations
\input{macros}

\begin{document}
\title[(Deep) Induction Rules for GADTs]{(Deep) Induction Rules for GADTs}
  \author{Patricia Johann, Enrico Ghiorzi}
  \email{johannp,ghiorzie@appstate.edu}

\begin{abstract} 
  Deep data types are those that are contructed from other data types,
  including, possibly, themselves.  In this case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction can be used to prove properties of
  nested types, including truly nested types, that cannot be proved
  via structural induction.  In this paper we show how to extend deep
  induction to GADTs that are not truly nested GADTs.  {\color{red}
    This opens the way to incorporating automatic generation of (deep)
    induction rules for them into proof assistants.}  We also show
  that the techniques developed in this paper do not suffice for
  extending deep induction to truly nested GADTs, {\color{red} so more
    sophisticated techniques are needed to derive deep induction rules
    for them.}
\end{abstract}

\maketitle

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep
ADTs~\cite{jp20}.\footnote{An ADT/nested type/GADT is {\em deep} if it
  is (possibly mutually inductively) defined in terms of other
  ADTs/nested types/GADTs (including, possibly, itself).}  The
following data type of rose trees, here coded in Agda and defined in
terms of the standard type $\mathsf{List}$ of lists (see
Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:
\[\begin{array}{l}
\mathsf{data\, Rose\, : Set \to Set\,where}\\
\mathsf{\;\;\;empty :\,\forall \{A : Set\} \to Rose\,A}\\
\mathsf{\;\;\;node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
\end{array}\]
The induction rule Coq automatically generates for (the analogous Coq
definition of) rose trees is
\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to}\\
\mathsf{\hspace*{0.15in}P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to}\\ 
\hspace*{0.15in}  \mathsf{\forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains.  With data types becoming ever more advanced, and with
deeply structured such types becoming increasingly ubiquitous in
formalizations, such an enhanced notion of induction is essential if
proof assistants are to be
%it is critically important that proof assistants
%%{\color{red} (e.g., Coq, Agda, Lean)}
%are
able to automatically generate genuinely useful induction rules for
data types that go
%well
beyond traditional ADTs. These
%Such data types
include not just deep ADTs, but also (truly\footnote{A truly nested
  type is a nested type that is defined over itself. The data type
  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
  concrete example.}) nested types~\cite{bm98},
%, such as $\mathsf{Rose}$ above, as well as truly nested types
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{dyb94}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to deliver induction rules
appropriate to nested types, including ADTs.  For the (deep) ADT of
rose trees, for example, it gives the following genuinely useful
induction rule:
\begin{equation}\label{eq:rose}
\begin{array}{l}
\mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
  \to}\\
\mathsf{\hspace*{0.15in}P\,empty \to}\\ 
\mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
  List\,(Rose\,A))\to Q\,a \to}\\
\mathsf{\hspace*{0.3in}List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
\mathsf{\hspace*{0.15in}\forall \,(x :
  Rose\,A) \to Rose^\land\,A\,Q\, x \to P\,x} 
\end{array}
\end{equation}
Here, $\mathsf{List^\land}$ (resp., $\mathsf{Rose^\land}$) lifts its
predicate argument $\mathsf{P}$ (resp., $\mathsf{Q}$) on data of type
$\mathsf{Rose\,A}$ (resp., $\mathsf{A}$) to a predicate on data of
type $\mathsf{List\,(Rose\,A)}$ (resp., $\mathsf{Rose\,A}$) asserting
that $\mathsf{P}$ (resp., $\mathsf{Q}$) holds for every element of its
list (resp., rose tree) argument.\footnote{Predicate liftings such as
  $\mathsf{List^\land}$ and $\mathsf{Rose^\land}$ can either be
  supplied as primitives or generated automatically from their
  associated data type definitions as described in
  Section~\ref{sec:ADTs-and-nesteds}. The predicate lifting for a
  container type like $\mathsf{List\,A}$ or $\mathsf{Rose\,A}$ simply
  traverses containers of that type and applies its predicate argument
  pointwise to the constituent data of type $\mathsf{A}$.  The ability
  to define predicate liftings for more general data types will be
  critical to deriving their deep induction rules in
  Section~\ref{sec:framework}.} Deep induction was also shown
in~\cite{jp20} to
% be the missing piece making it possible to
%be the theretofore missing component needed to
% be the essential ingredient making it possible to
deliver the first-ever induction rules --- structural or otherwise ---
for the $\mathsf{Bush}$ data type~\cite{bm98} and other truly nested
types. Deep induction for ADTs and (truly) nested types is reviewed in
Section~\ref{sec:ADTs-and-nesteds}.

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of data whose types are pair-structured, rather
than sequences of arbitrary type, as does $\mathsf{const}$.

If one or more of the data constructors for a GADT $\mathsf{G}$ return
structured instances of $\mathsf{G}$, then the GADT will have two
distinct, but equally natural, semantics: a functorial semantics
interpreting it as a left Kan extension~\cite{mac71}, and a parametric
semantics interpreting it as the interpretation of its Church
encoding~\cite{atk12,vw10}. As detailed in~\cite{jgj21}, a key
difference in the two semantics is that the former views GADTs as
their {\em functorial completions}~\cite{jp19}, and thus as containing
more data than just those expressible in syntax. By contrast, the
latter views them as what might be called {\em syntax-only}
GADTs. Happily, these two views of GADTs coincide for those GADTs that
are ADTs or (other, including truly) nested types.  However, both they
and their attendant properties differ greatly for proper GADTs. In
fact, the functorial and parametric semantics for proper GADTs are
sufficiently disparate that, by contrast with the semantics
customarily given for ADTs and nested
types~\cite{bfss90,gjfor15,jgj21f}, it is not at all clear how to
define a functorial parametric semantics for GADTs~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so GADTs viewed as their functorial completions
do not obviously support induction rules.  This makes sense
intuitively: induction is a syntactic proof technique, so it may not
be possible to use it to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs. Indeed, this
paper gives the first-ever deep induction rules
%--- deep or otherwise ---
for proper GADTs.
%It also shows how the structural induction rule for
%a GADT can be derived as an instance of its deep induction rule.
But it actually delivers far more: it gives a {\em general framework}
for deriving deep induction rules for GADTs that can be instantiated
to particular such data type of interest. This framework can serve as
a basis for extending modern proof assistants' automatic generation of
structural induction rules for ADTs to automatic generation of deep
induction rules for GADTs. In addition, as for ADTs and nested types,
the structural induction rule for any GADT can be recovered from its
deep induction rule by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

@@

A structural induction rule for a data type allows us to prove that if
a predicate holds for every element inductively produced by the data
type's constructors then it holds for every element of the data type.
In this paper, we are interested in induction rules for proof-relevant
predicates.  A proof-relevant predicate on $\mathsf{A : Set}$ is a
function $\mathsf{P\,:\,A \to Set}$ mapping each $\mathsf{a : A}$ to
the set of proofs that $\mathsf{P\,a}$ holds.  For example, the
structural induction rule for the list type
\begin{equation*}\label{eq:list}
\begin{array}{l}
\mathsf{data\ List : Set \to Set\ where}\\
\mathsf{\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
\mathsf{\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
\end{array}
\end{equation*}
is
\begin{equation*}
  \begin{array}{l}
  \mathsf{\forall (A : Set) (P : List\,A \to Set) \to}\\
  \mathsf{\hspace*{0.15in}P\,nil \to}\\
  \mathsf{\hspace*{0.15in}\big( \forall (a : A) (as: List\,A) \to P\,as
    \to P\,(cons\,a\,as)\big) \to}\\
  \mathsf{\hspace*{0.15in}\forall (as : List\,A) \to P\, as}
\end{array}
\end{equation*}
As in Coq's induction rule for rose trees, the data inside a structure
of type $\mathsf{List}$ is treated monolithically (i.e., is ignored)
by this structural induction rule.  By contrast, the deep induction
rule for lists is parameterized over a custom predicate $\mathsf{Q}$
on $\mathsf{A}$. For $\mathsf{List^\wedge}$ as described in the
introduction the deep induction rule for lists is
\[\begin{array}{l}
\mathsf{\forall (A : Set) (P : List\, A \to Set) (Q : A \to Set)
  \to}\\
\mathsf{\hspace*{0.15in}P\,nil \to}\\
\mathsf{\hspace*{0.15in}\big( \forall (a : A) (as: List\,A) \to Q\,a \to P\,as
\to P\,(cons\,a\,as)\big) \to} \\ 
\mathsf{\hspace*{0.15in}\forall (as : List\,A) \to List^{\wedge}\,A\,Q\,as
  \to P\,as } 
\end{array}\]

Structural induction can be extended to nested types, such as the
following type of perfect trees~\cite{bm98}:
\begin{equation*}\label{eq:ptree}
\begin{array}{l}
\mathsf{data\ PTree : Set \to Set\ where}\\
\mathsf{\;\;pleaf\,\,\; :\,\forall \{A : Set\} \to A \to PTree\,A}\\
\mathsf{\;\;pnode\, :\,\forall \{A : Set\} \to PTree\,(A \times A) \to PTree\,A} 
\end{array}
\end{equation*}
Perfect trees can be thought of as lists constrained to have lengths
that are powers of 2. In the above code, the constructor
$\mathsf{pnode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
construct data of type $\mathsf{PTree\,A}$. Thus, it is clear that the
instances of $\mathsf{PTree}$ at various indices cannot be defined
independently, and that the entire inductive family of types must
therefore be defined at once. This intertwinedness of the instances of
nested types is reflected in their structural induction rules, which,
as explained in~\cite{jp20}, must necessarily involve polymorphic
predicates rather than the monomorphic predicates appearing in
structural induction rules for ADTs. The structural induction rule for
perfect trees, for example, is
\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set) \to}\\
\mathsf{\hspace*{0.15in}\big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to}\\
\mathsf{\hspace*{0.15in}\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to}\\
\mathsf{\hspace*{0.3in}P\,(A \times A)\,pp \to P\,A\,(pnode\,pp)\big) \to}\\
\mathsf{\hspace*{0.15in}\forall (A : Set) (p : PTree\,A) \to P\,A\,p} 
\end{array}\]

\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
      -0.2in 0in, fbox, center}

{\small
\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A
  \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
  Q\,a \to P\,A\,Q\,(pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
  \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
  P\,A\,Q\,(pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
  : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
  P\,A\,Q\,p }\\
 \\
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
\to \big( \forall (A : Set)\,(Q : A \to Set) \to P\,A\,Q\,bnil \big) \to} \\ 
\quad\mathsf{\big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
  Bush\,(Bush\,A)) \to Q\,a \to
  P\,(Bush\,A)\,(P\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
  \big) \to} \\ 
\quad\mathsf{\forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
  Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
\end{array}\]}

@@

\caption{Deep induction rules for perfect trees and
  bushes}\label{fig:ptree-and-bush}
\end{adjustbox}
\end{figure*}

The deep induction rule for perfect trees similarly uses polymorphic
predicates but otherwise follows the familiar pattern. It is given
by the first expression in Figure~\ref{fig:ptree-and-bush}. There,
$\mathsf{Pair^{\wedge} : \forall (A\; B: Set) \to (A \to Set) \to (B
  \to Set)\; \to\;}$ $\mathsf{A \times B \to}$ $\mathsf{Set}$ lifts predicates
$\mathsf{Q_A}$ on data of type $\mathsf{A}$ and $\mathsf{Q_B}$ on data
of type $\mathsf{B}$ to a predicate on pairs of type $\mathsf{A \times
  B}$ in such a way that $\mathsf{Pair^{\wedge}\,A\,B\,Q_A\,Q_B\,(a,b)
  = Q_A\,a \times Q_B\,b}$. Similarly, $\mathsf{PTree^{\wedge}}$
$\mathsf{\,: \,\forall (A : Set) \to (A \to Set) \to PTree\,A \to Set}$
lifts a predicate $\mathsf{Q}$ on data of type $\mathsf{A}$ to a
predicate on data of type $\mathsf{PTree\,A}$ asserting that
$\mathsf{Q}$ holds for every element of type $\mathsf{A}$ contained in
its perfect tree argument.\looseness=-1

Using deep induction we can extend structural induction to {\em truly
  nested types}, i.e., to nested types whose recursive occurrences
appear below themselves. The quintessential example of such a type is
that of bushes\footnote{To define truly nested types in Agda we must
  use the $\mathsf{NO\_POSITIVITY\_CHECK}$ flag, and to define
  functions over them we must use the $\mathsf{TERMINATING}$ flag.
  {\color{red}(Similar flags are required in Coq.)} Although as
  programmers we know from the metatheory in~\cite{jp19} that
  $\mathsf{Bush}$ is well-defined and the functions we define over
  them terminate, the flags are necessary because Agda fails to infer
  these facts. Analogous comments apply at several places
  below.}\cite{bm98}:
\begin{equation*}\label{eq:bush}
\begin{array}{l}
\mathsf{data\ Bush : Set \to Set\ where}\\
\mathsf{\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
\mathsf{\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
\end{array}
\end{equation*}
Even defining a structural induction rule for bushes requires that we
be able to lift the rule's polymorphic predicate argument to
$\mathsf{Bush}$ itself. This observation was, in fact, the original
motivation for the development of deep induction in~\cite{jp20}. The
deep induction rule for bushes is given by the second expression in
Figure~\ref{fig:ptree-and-bush}, where $\mathsf{Bush^{\wedge} :
  \forall (A : Set) \to (A \to Set)}$ $\mathsf{\to Bush\,A \to Set}$
is the following lifting of a predicate $\mathsf{Q}$ on data of type
$\mathsf{A}$ to a predicate on data of type $\mathsf{Bush\,A}$
asserting that $\mathsf{Q}$ holds for every element of type
$\mathsf{A}$ contained in its argument bush:
\begin{equation}
\begin{array}{l}
\mathsf{Bush^{\wedge}\,A\,Q\,bnil} \hspace*{0.49in} = \; \mathsf{\top} \\
\mathsf{Bush^{\wedge}\,A\,Q\,(bcons\,a\,bb)} \; = \;\\
\hspace*{0.3in}\mathsf{Q\,a \times Bush^{\wedge}\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb} 
\end{array}
\end{equation}

Under some circumstances deep induction can be mimicked by
hand-threading applications of structural induction through the layers
of data comprising a deep data type. But this is not the case if,
e.g., one or more of the custom predicates in the data type's deep
induction rule is not the characteristic function of an inductive data
type.

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.  For every
GADT in this paper, we require that the instance of the return type
for each of its data constructors is a polynomial in that
constructor's type arguments.\looseness=-1

GADTs are used in precisely those situations in which different
behaviors at different instances of data types are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then using pattern
matching to force the desired type refinement. For example, the {\em
  equality} GADT
\begin{equation}\label{eq:equal}
\begin{array}{l}
\mathsf{data\ Equal : Set \to Set \to Set\ where}\\
\mathsf{\;\;refl :\, \forall \{A : Set\} \to Equal\,A\,A}
\end{array}
\end{equation}
is parameterized by two type indices, but it is only possible to
construct data elements of type $\mathsf{Equal\,A\,B}$ if $\mathsf{A}$
and $\mathsf{B}$ are instantiated at the same type. If the types
$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
type $\mathsf{Equal\,A\,B}$ contains the single data element
$\mathsf{refl}$. It contains no data elements otherwise.

\begin{adjustbox}{varwidth=7.3in, max width=7.1in, margin=-0.0in 0in
      -0.2in 0in, fbox, center}
